Metagenomic Barcoding of Pollen Loads Offers Insights on the Foraging

### 2022-12-11 script/manuscript_MEE.Rmd
most of ST edits and suggestions incorporated into manuscript

---
title: Metagenomic Barcoding of Pollen Loads Offers Insights on the Foraging
  Patterns of Queen Bumble Bees
author: |
  Reed Clark Benkendorf^[Correspondence: reedbenkendorf2021@u.northwestern.edu] $^1$$^,$$^2$, Jane E. Ogilvie $^3$, Emily J. Woodworth $^1$$^,$$^2$, Sophie Taddeo $^1$$^,$$^2$, Paul J. CaraDonna $^1$$^,$$^2$$^,$$^3$, Jeremie B. Fant $^1$$^,$$^2$ | $^1$Chicago Botanic Garden, 1000 Lake Cook Road, Glencoe, Illinois 60022 USA,  $^2$Plant Biology and Conservation, Northwestern University, Evanston, IL 60208, USA, $^3$ Rocky Mountain Biological Laboratory, P.O. Box 519, Crested Butte, CO 81224, USA
abstract: |
  .    
  1) DNA Barcoding has been remarkably successful in nearly all kingdoms of life and has allowed for the rapid analysis of ecological assemblies. Successful DNA barcoding in the plant kingdom has been more difficult than other kingdoms. Due to this understanding plants in ecological contexts and understandings of their synecology in some instances may begin to lag behind other kingdoms.  
  2) Here we utilize hyb-seq, museum studies, and species distribution modelling, to detect the plant species present in pollen loads collected from Queen Bumble Bees.  
  3) We show that hyb-seq using the Angiosperms 353 probes offers significant promise to metagenomic approaches in real world scenarios.  
  4) We conclude that these probes offer promise for the identification of plant tissue in both single sample, and metasample contexts.

@@

  Recently barcoding, and metabarcoding, have shown considerable promise in all Kingdoms of life. For example …. . With plants the identification of members of certain clades has been quite successful, whereas with others results have been elusive (@liu2014identification, @china2011comparative), while most applications laying along this spectrum (@li2015plant, @kress2007two, @cbol2009dna). Particular challenges with the utilization of high-copy number sequences are associated with their rates of divergence, gene tree conflict, and hybridization (@coissac2016barcodes, @fazekas2009plant) Herein we have resolved major components of the problems of identifying plant material without diagnostic morphological character states using the Angiosperms353 (A353) Hyb-Seq probes (@johnson2019universal), and custom species sequence databases derived via species distribution modelling, and temporal filtering.

  Our foundation for increasing the quality of metabarcoding results in plants is reducing the number of possible plant species candidates by generating user selected sequence databases relevant to the spatial extent of the study region. While there are numerous possible approaches for this process, we achieve the selection of possible plant candidate species using digital collections gleaned from herbaria, survey work, and citizen science (e.g. iNaturalist), from a domain exceeding the study area. To these candidate species, modelling approaches - such as logistic regression, may be used to identify distances under which taxa warrant further exploration. To these candidate species, we generate species distribution models (SDM's), which indicate the probability of suitable habitat in a domain, and base the inclusion of these taxa, or representative congeners, upon these results. This approach has the additional benefit of greatly reducing the size of a sequence database, which allows for the usage of genomic size data on personal computers. Moreover, as most next-generation sequence data is deposited as raw-sequence reads, from a processing perspective, it is essential to reduce the candidate species via an approach as such.

  Currently the largest plant systematic endeavor ever undertaken, the Kew Plant and Fungal Tree of Life (PAFTOL), is approaching completion (@baker2021PAFTOL). This dataset will contain Hyb-Seq data from at least one species representing each genus in the plant kingdom using the popular A353 probes (@baker2021exploring), resulting in over 14,000 represented species. These publicly available data serve to provide a taxonomically comprehensive backbone for plant metabarcoding. Data from the 10kP project, which seeks to develop reference genomes from a phylogenetically diverse suite of plants will contribute many more records upon it's intended completion, now slated to be by 2030, similar projects which seek to sequence high amounts of genomes in regions e.g. the 'Darwin Tree of Life' are being undertaken which will contribute data for applicable to enormous spatial domains (@cheng2018tenkp, @darwin2022project, @lewin2022biogenome).

  Considerable amounts of species interactions are expressed along time (@caradonna2021seeing). For the tropics the flowering periods of many plant species display high seasonality, and given the elevated rates of species richness relative to the temperate, this axis may provide an essential filter for identifying material in many metagenomic samples (@janzen1967synchronization, @Newstrom1994ANC). While many pollination interactions are formed and dissolved along the temporal axis in the temperate regions (@caradonna2017interaction), the overall shorter extent of the active growing season in these systems results in the presence of few to any natural breaks in these systems which subjugates the utility of these to perform as filters of post-processing results, rather than distinct species assemblage for database generation. Nonetheless, we work through a process which seems applicable to the tropics to utilize the temporal dimension for classifying sequencing results.

@@

Observations and sample collection was conducted at The Rocky Mountain Biological Laboratory (RMBL; 38°57.5" N, 106°59.3" W (WGS 84), 2900 m.a.s.l.), Gunnison County, Colorado, USA (*APPENDIX 1* for site information). Pollinator observations of *Bombus* Latreille spp. (Apidae Latreille) were conducted from June - August of 2015. Observations of *Bombus* foraging took place for one hour at each field site in three 100m transects, where all flowers were also counted and place into abundance bins. Corbiculae loads were,  non-lethally, collected once from all Queen individuals encountered. The six study sites are in areas characterised by high-montane/subalpine Parkland vegetation communities.

@@

To develop an ecologically relevant list of vascular plant species, with expected biotic pollination, which may be present at the study sites all records adjacent to the field site were downloaded from the Botanical Information and Ecology Network 'BIEN' (@bien2022), and these taxa had Species Distribution Models (SDMs) generated to predict their suitability. The predicted plant species served as a reference for which species to include in the genomic sequence databases.

In order to minimise the number of species for which SDM’s were to be generated, BIEN was queried at a distance of up to 100km from our field site and all plant species records were downloaded. In order to emulate the perceived stochasticity of collections, this dataset was bootstrap re-sampled 250 times, with 90% of samples selected, to create a testing dataset. The median of the logistic regression assessing the probability of occurrence of a species record as a function of distance from the study area was used as a threshold distance, under which, to include species as candidates for distribution modelling.

@@

Species had all records from BIEN within a 50km border of the Omernik level 3 ecoregion which the site is located in *(No. 21 "Southern Rockies")*, downloaded (n = 23,919) (@omernik1987ecoregions). These records were copied into two, initially identical, sets, one for generating machine learning models (Random Forest, and Boosted Regression Tree's), and the other for Generalised Linear (GLM) and Generalized Additive Models (GAM). The set for generating GLM and GAM records was thinned to reduce spatial autocorrelation in the dataset, as measured by Morans Index (@moran1950notes, @spdep). To both datasets an additional 4029 plots collected from a random stratification of 19% of the land cover in the area of analysis were searched to create true absences (BLM CITATION - need appropriate format for journal). To achieve a larger absence dataset 1000 pseudo-absence records were generated for each taxon, each of which was greater than 10km from an occurrence record. For ML models, these pseudo-absences were reduced so that the ratio of presence to absence records were balanced. To achieve this, absence records inside of 10% of the mean sample value of the presence records, for any predictor were removed; the required number of absence records were then randomly sampled.

Species abiotic niche predictors were 26 variables at 30m resolution, six related to climate,  five soil, four topographic, four related to cloud cover, with the remaining reflecting assorted abiotic parameters (@wilson2016remotely, @wang2016locally, @hengl2017soilgrids250m, @robinson2014earthenv) (*APPENDIX 6*). For linear regression models these predictors underwent both vifstep (theta = 10, max observations = 12,500) and vifcor (theta = 0.7, max observations = 12,500), and collinear features were removed leaving 16 variables (@usdm2014).

Modelling: Random Forest and Boosted Regression Trees, were sub sampled with 30% test and two replicates each before weighted ensemble based on True Skill Statistics (tss) (@sdmPackage). Generalised linear models (GLM) and Generalised additive models (GAM) with 30% sub sampling and three replicates each were also ensembled using the tss (@sdmPackage). The results of these models were extracted to a polygon feature derived from a minimum-spanning tree which encompasses the study area, and species from either ensemble with greater than 50% habitat suitability were considered present for further purposes (@prim1957minimum).

535 species were modelled using Generalized Linear Models and Generalized Additive Models. 534 species were modelled using Random Forest and Boosted Regression Trees.  To evaluate the accuracy of the species distribution models, additional presence records from GBIF (n = 61,789), and AIM (n = 12,730) were used as test and training sets (n = 74,519) for logistic regression (CITE AIM AND GBIF). Additional novel absence records were generated from the AIM dataset to create a dataset where each species has balanced presence and absences. 11 or more paired presence and absence records were required for this testing, resulting in 334 species being included in the logistic regression (Mdn = 110.0, $\bar{x}$ = 223.1, max = 1568 record pairs used) with a 70% test split (@caret).

@@

Using 5 years of observational data on *Bombus* Queen Bee foraging at these studies sites, we identified the plant taxa most frequently visited by Queens across all years. We sequenced the 12 most commonly visited taxa twice using samples from one site within the Gunnison River Drainage and one individual from another population. In addition, for any of these 12 focal species which did not have a congener pair in this filtered sample, we included a congener - or a species from a closely related genus to serve as an outgroup. We also sequenced another 15 abundant taxa commonly visited by *Bombus* workers, based on the aforementioned data set (*APPENDIX 4*).

@@

To develop a reference library of  pollen grains which may be present in corbiculae loads, an image reference collection of fuchsin-jelly stained (@beattie1971technique) slides was assembled from slides previously prepared by the authors (n = 21), and other researchers (n = 38) (@brosi2013single). Using 5 years of observational data on *Bombus* Queen Bee foraging at these studies sites (Ogilvie unpublished), as well as the Vascular Plant Checklist (@fraser2007vpc), an additional 62 voucher slides for species were prepared and imaged at 400x (Leica DMLB, Leica MC170 HD Camera, Leica Application Suite V. 4.13.0) from non accessioned herbarium collections to supplement the number of species and clades covered (Appendix 3).  
  In order to determine which plant taxa were distinguishable via light microscopy, and to develop a dichotomous key to pollen morphotypes, Divisive Hierarchical Clustering techniques were used. Ten readily discernible categorical traits were collected from each specimen in the image collection. These traits were transformed using Gower distances, and clustered using Divisive Hierarchical clustering techniques (@cluster2022). Using the cluster dendrogram, elbow plot, and heatmaps (@fpc2022), of these results morphological groups of pollen which could not be resolved via microscopy were delineated, and a dichotomous key was prepared (APPENDIX NO.). This key was then used to identify the pollen grains sampled from corbiculae loads to morphotypes in a consistent manner. 
  To prepare the pollen slides from corbiculae, all corbiculae loads were broken apart and rolled using dissection needlepoints to increase heterogeneity of samples. *Cerca* 0.5mm^2 of pollen was placed onto a ~4mm^2 fuchsin jelly cube (@beattie1971technique) atop a graticulated microscope slide, with 20 transects and 20 rows (400 quadrants) (EMS, Hartfield, PA). The jelly was melted, with stirring, until pollen grains were homogeneously spread across the microscope slide. Slides were sealed with Canada Balsam (Rublev Colours, Willits, CA) followed by sealing with nail polish; all samples are noted in *APPENDIX 3*.
  To identify the pollen present in corbiculae loads, light microscopy at 400x (Zeiss Axioscope A1) was used. In initial sampling in three transects, each pollen grain was identified to morphotype and counted; an additional two transects were scanned for morphotypes unique to that slide, if either transect contained an unique morphotype than all grains in that transect were also identified and counted. Subsequent to the first round of sampling, non-parametric species richness rarefaction curves (@vegans2022), and non-parametric species diversity rarefaction curves were used to assess the completeness of sampling (@inextArticle, @inextPackage). Slides not approaching the asymptote of the rarefaction curve were then re-sampled, and analysed iteratively for up to a total of seven transects *APPENDIX 2*.

@@

To estimate the duration of dates in which plant species were flowering weibull estimates of several phenological parameters all spatially modelled taxa were developed (@belitz2020accuracy, @pearse2017statistical). Only BIEN records which occurred in the Omernik Level 4 Ecoregions within 15km of the study area (n = 5, or conditionally 6 if enough records not be found in the nearest 5), and which were from herbarium records were included. To remove temporally irrelevant herbarium records, i.e. material collected during times which flowering is impossible at the study area due to snow cover, the SnowUS dataset (@iler2021conceptual, @tran2019cloud) from 2000-2017 was analyzed for the first three days of contiguous snow absence, and the first three days of contiguous snow cover in Fall. Herbarium records after the 3rd quantile for melt, and the 1st quantile for snow cover of these metrics were removed. Species with > 10 records had their weibull distributions generated for the date when 10% of individuals had begun flowering, when 50% were flowering, and when 90% of individuals had flowered.

@@

Across the entire spatial domain of modelling all ensembled models (n = 968) had an accuracy of 0.84 (95% CI 0.8356 - 0.8443), kappa 0.68, p-value < 0.001, sensitivity = 0.80, specificity = 0.87.

@@

**ACKNOWLEDGMENTS:** Nyree Zerega for assistance obtaining herbaria loans and accessioning our collections at CHIC. Pat Herendeen for assistance with virtually all aspects of preparing pollen vouchers and the identification process. Hilary Noble, Zoe Diaz-Martinez, Angela McDonnell, & Elena Loke for assistance with genomic library preparation. We thank the curators at the following herbaria for supplying tissue: Ben Legler at Stillinger (ID), Charles (Rick) Williams at Ray J. Davis (IDS), (B)Ernie Nelson at Rocky Mountain (RM). We thank the original collectors of these specimens: D. Knoke, L. Brummer, J. Boyd, C. Davidson, I. Gilman, M. Kirkpatrick, S. McCauley, J. Smith, K. Taylor, & C. Williams. David Giblin for sharing relevant sections of an advanced draft of FNA V. 15. The Bureau of Land Management is thanked as many plant specimens were collected by R.C.B as a partner or contractor to the agency, Sarah Burnett and Lauren Price are thanked for sharing AIM data. Sanda and NEB are gratefully acknowledged for technical support and generously sharing samples.

@@

>>>>

---
title: Metagenomic Barcoding of Pollen Loads Offers Insights on the Foraging
  Patterns of Queen Bumble Bees
author: |
  Reed Clark Benkendorf^[Correspondence: reedbenkendorf2021@u.northwestern.edu] $^1$$^,$$^2$, Jane E. Ogilvie $^3$, Emily J. Woodworth $^1$$^,$$^2$, Sophie Taddeo $^1$$^,$$^2$, Paul J. CaraDonna $^1$$^,$$^2$$^,$$^3$, Jeremie B. Fant $^1$$^,$$^2$ | $^1$Chicago Botanic Garden, 1000 Lake Cook Road, Glencoe, Illinois 60022 USA,  $^2$Plant Biology and Conservation, Northwestern University, Evanston, IL 60208, USA, $^3$ Rocky Mountain Biological Laboratory, P.O. Box 519, Crested Butte, CO 81224, USA
abstract: |
  .    
  1) DNA Barcoding has been remarkably successful in nearly all kingdoms of life and has allowed for the rapid analysis of ecological assemblies. Successful DNA barcoding in the plant kingdom has been more difficult than other kingdoms. Due to this understanding plants in ecological contexts and understandings of their synecology in some instances may begin to lag behind other kingdoms.
  2) Here we utilize hyb-seq, museum studies, and species distribution modelling, to detect the plant species present in pollen loads collected from Queen Bumble Bees.
  3) By utilizing Species distribution modelling we allow for one to process hyb-seq data, create user specified sequence databases which may use MORE ACCURATE alignment algorithms on personal computers over realistic time periods. 
  4) We show that hyb-seq using the Angiosperms 353 probes, which are currently being used in the largest ever plant systematic endeavor, offers significant promise to metagenomic approaches in real world scenarios.
  5) We conclude that these probes offer promise for the identification of plant tissue in both single sample, and metasample contexts.

@@

  Recently barcoding (the identification of a sample from a single organism – e.g. a piece of leaf), and metabarcoding (the identification of a sample containing a mix of organisms – e.g. soil), have shown considerable promise in all Kingdoms of life (@ruppert2019past). With plants the identification of members of certain clades using barcoding has been quite successful (REF), whereas with others results have been elusive (@liu2014identification, @china2011comparative, @coissac2012bioinformatic), however metabarcoding incurs additional challenges to those which  exist for the currently implemented barcodes (@li2015plant, @kress2007two, @cbol2009dna, @coissac2012bioinformatic). Particular challenges with barcodes include the utilization of high-copy number sequences are associated with their rates of divergence, gene tree conflict, and hybridization (@coissac2016barcodes, @fazekas2009plant). Particular challenges with the utilization of high-copy number sequences are associated with their rates of divergence, gene tree conflict, and hybridization (@coissac2016barcodes).

  Currently the largest plant systematic endeavor ever undertaken, the Kew Plant and Fungal Tree of Life (PAFTOL), is approaching completion (@baker2021PAFTOL). This dataset will contain Hyb-Seq data from at least one species representing each genus in the plant kingdom using the popular A353 probes (@baker2021PAFTOL), resulting in over 14,000 represented species. These publicly available data serve to provide a taxonomically comprehensive backbone for plant metabarcoding. Data from the 10kP project, which seeks to develop reference genomes from a phylogenetically diverse suite of plants will contribute many more records upon it’s intended completion, now slated to be by 2030, similar projects which seek to sequence high amounts of genomes in regions e.g. the ‘Darwin Tree of Life’ are being undertaken which will contribute data  applicable to enormous spatial domains (@cheng2018tenkp, @darwin2022project, @lewin2022biogenome). These data will promote the ability to apply metabarcoding to resolve a diversity of questions relevant to theoretical and applied ecology (cite). However, the application of metabarcoding still face challenges relating to the enormity of the genomic datasets and the computational power required to process sequence data. Herein we have resolved major components of the problems of identifying plant material without diagnostic morphological character states using the Angiosperms353 (A353) Hyb-Seq probes (@johnson2019universal), and custom species sequence databases derived via species distribution modelling, and temporal filtering.

  To increase the quality of metabarcoding results in plants, we suggest reducing the number of possible plant species candidates by generating user selected sequence databases relevant to the the study region and its ecological characteristics (CITE !?). To achieve this goal, we first create a list of candidate species using digital collections gleaned from herbaria, survey work, and citizen science (e.g. iNaturalist), from a region exceeding the study area. To these candidate species, modelling approaches - such as logistic regression, may be used to identify taxa which warrant further exploration e.g. modelling to determine their possibility of presence in metabarcoding samples. We then use species distribution models to create potential distribution maps for the candidate species to limit the impact of spatial and taxonomic biases in the species list and account for spatial variations in niche availability throughout the study area. Species distribution models (SDM’s) examine the ecological conditions associated with known occurrence of a species to identify where else in the study area might suitable habitats be found. This approach has the additional benefit of greatly reducing the size of a sequence database, which allows for the usage of genomic size data on personal computers. This can also significantly reduce processing time, particularly as as most next-generation sequence data is deposited as raw-sequence reads.

  Considerable amounts of species interactions vary along time (@caradonna2021seeing). For the tropics, contrasts in the flowering periods of many plant species, can provide an additional filter for identifying material in many metagenomic samples (@janzen1967synchronization, @Newstrom1994ANC). In temperate regions, pollination interactions also vary temporally (@caradonna2017interaction), the overall shorter extent of the active growing season in these systems results in the presence of few to any natural breaks in these systems which subjugates the utility of these to perform as filters of post-processing results, rather than distinct species assemblage for database generation. Nonetheless, we work through a process which seems applicable to the tropics to utilize the temporal dimension for classifying sequencing results.

@@

Observations and sample collection was conducted at The Rocky Mountain Biological Laboratory (RMBL; 38°57.5” N, 106°59.3” W (WGS 84), 2900 m.a.s.l.), Colorado, USA (APPENDIX 1 for site information). Pollinator observations of Bombus Latreille spp. (Apidae Latreille) were conducted from June - August of 2015 in six study sites characterized by high-montane/subalpine Parkland vegetation communities. Observations of Bombus foraging took place for one hour at each field site in three 100m transects, where all flowers were also counted and placed into abundance bins. Corbiculae loads were, non-lethally, collected once from all Queen individuals encountered.

@@

We downloaded from the Botanical Information and Ecology Network ‘BIEN’ (@bien2022) all records adjacent to the field sites to develop an ecologically relevant list of vascular plant species, with expected biotic pollination, which may be present at the study area . We then generated Species Distribution Models (SDMs) to predict their distribution throughout the study area. These maps of potential distribution served as a reference to reduce the list of species to include in the genomic sequence databases.

In order to minimise the number of species for which SDM’s were to be generated, BIEN was queried at a distance of up to 100km from our study area and all plant species records were downloaded. **_In order to emulate the stochasticity of botanical collecting, this dataset was bootstrap re-sampled 250 times, with 90% of samples selected, to create a testing dataset_**. The median of the logistic regression assessing the probability of occurrence of a species record as a function of distance from the study area was used as a threshold distance, under which, to include species as candidates for distribution modelling.

@@

We used all occurrence records from BIEN (n = 23,919) within a 50km border of the Omernik level 3 ecoregion, which includes the study area *(No. 21 “Southern Rockies”)* to construct the species distribution model  (@omernik1987ecoregions). These records were copied into two, initially identical, sets, one for generating machine learning models (ML; Random Forest, and Boosted Regression Tree’s), and the other for Generalised Linear (GLM) and Generalized Additive Models (GAM) (@barbet2012selecting). **Ensembled predictions have been shown to outperform their constituent models, on average, and to reduce the ecological signal to the analytical noise of individual runs (@araujo2007ensemble). No single method of producing SDMs has been shown to universally outperform others when faced with a large and diverse number of applications, in our case a great number of species with differing biologies and ecologies (@elith2006novel, @qiao2015no). In the spirit of these findings, multiple families of models, which can be generated together as they have similar requirements regarding the number and ratios of Presence to Absence records were ensembled together (@barbet2012selecting).**

We then generated 4,029 absence points , locations where the focal taxon is anticipated missing, through a random stratification of 19% of the land cover in the area and included them in (BLM CITATION - need appropriate format for journal). To achieve a larger absence dataset, we generated 1,000 pseudo-absence records for each taxon by randomly selecting coordinates located at least 10km away from an occurrence record. For ML models, these pseudo-absences were reduced so that the ratio of presence to absence records were balanced (@barbet2012selecting). To achieve this, we removed absence records inside of 10% of the mean sample value of the presence records; the required number of absence records were then randomly sampled.

We used 26 environmental variables at 30m resolution to predict the potential distribution of each species, six related to climate, five soil, four topographic, four related to cloud cover, with the remaining reflecting assorted abiotic parameters (@wilson2016remotely, @wang2016locally, @hengl2017soilgrids250m, @robinson2014earthenv) (*APPENDIX 6*). **These publicly available datasets, were selected as they ...** . For linear regression models these predictors underwent both *vifstep* (theta = 10, max observations = 12,500) and *vifcor* (theta = 0.7, max observations = 12,500) to detect highly correlated variables, and collinear features were removed leaving 16 variables (@usdm2014).

Modelling: Random Forest and Boosted Regression Trees, were sub sampled with 30% test and two replicates each before weighted ensemble based on True Skill Statistics (tss) (@sdmPackage). Generalised linear models (GLM) and Generalised additive models (GAM) with 30% sub sampling and three replicates each were also ensembled using the tss (@sdmPackage, @). TSS was chosen as the ensemble criterion as it has been shown to work across a range of species occurrences prevalences (@allouche2006assessing). The results of these models were extracted on a cell-by-cell basis to a polygon feature derived from a minimum-spanning tree which encompasses the study sites, and species from either ensemble with greater than 50% mean habitat suitability across all cells were considered present for further purposes (@prim1957minimum).

535 species were modelled using Generalized Linear Models and Generalized Additive Models. 534 species were modelled using Random Forest and Boosted Regression Trees.  To evaluate the accuracy of the species distribution models, additional presence records from GBIF (n = 61,789), and AIM (n = 12,730) were used as test and training sets (n = 74,519) for logistic regression (CITE AIM AND @gbifDL2021sdms). Additional novel absence records were generated from the AIM dataset to create a dataset where each species has balanced presence and absences. 11 or more paired presence and absence records were required for this testing, resulting in 334 species being included in the logistic regression (Mdn = 110.0, $\bar{x}$ = 223.1, max = 1568 record pairs used) with a 70% test split (@caret).

@@

Using five years of observational data on *Bombus* Queen Bee foraging at these studies sites, we identified the plant taxa most frequently visited by Queens across all years. We sequenced the 12 most commonly visited taxa twice using samples from one site within the Gunnison River Drainage and one individual from another population. In addition, for any of these 12 focal species which did not have a congener pair in this filtered sample, we included a congener - or a species from a closely related genus to serve as an outgroup. We also sequenced another 15 abundant taxa commonly visited by *Bombus* workers, based on the aforementioned data set (*APPENDIX 4*).

@@

To develop a reference library of  pollen grains which may be present in corbiculae loads, an image reference collection of fuchsin-jelly stained (@beattie1971technique) slides was assembled from slides previously prepared by the authors (n = 21), and other researchers (n = 38) (@brosi2013single). Using five years of observational data on *Bombus* Queen Bee foraging at these studies sites (Ogilvie unpublished), as well as the Vascular Plant Checklist (@fraser2007vpc), an additional 62 voucher slides for species were prepared and imaged at 400x (Leica DMLB, Leica MC170 HD Camera, Leica Application Suite V. 4.13.0) from non accessioned herbarium collections to supplement the number of species and clades covered (Appendix 3).  
  We used Divisive Hierarchical Clustering techniques to determine which plant taxa were distinguishable via light microscopy, and to develop a dichotomous key to pollen morphotypes. Ten readily discernible categorical traits were collected from each specimen in the image collection. These traits were transformed using Gower distances, and clustered using Divisive Hierarchical clustering techniques (@cluster2022). Using the cluster dendrogram, elbow plot, and heatmaps (@fpc2022), of these results morphological groups of pollen which could not be resolved via microscopy were delineated, and a dichotomous key was prepared (APPENDIX NO.). This key was then used to identify the pollen grains sampled from corbiculae loads to morphotypes in a consistent manner. 
  To prepare the pollen slides from corbiculae, all corbiculae loads were broken apart and rolled using dissection needlepoints to increase heterogeneity of samples. *Cerca* 0.5mm^2 of pollen was placed onto a ~4mm^2 fuchsin jelly cube (@beattie1971technique) atop a graticulated microscope slide, with 20 transects and 20 rows (400 quadrants) (EMS, Hartfield, PA). The jelly was melted, with stirring, until pollen grains were homogeneously spread across the microscope slide. Slides were sealed with Canada Balsam (Rublev Colours, Willits, CA) followed by sealing with nail polish; all samples are noted in *APPENDIX 3*.
  To identify the pollen present in corbiculae loads, light microscopy at 400x (Zeiss Axioscope A1) was used. In initial sampling in three transects, each pollen grain was identified to morphotype and counted; an additional two transects were scanned for morphotypes unique to that slide, if either transect contained an unique morphotype than all grains in that transect were also identified and counted. Subsequent to the first round of sampling, non-parametric species richness rarefaction curves (@vegans2022), and non-parametric species diversity rarefaction curves were used to assess the completeness of sampling (@inextArticle, @inextPackage). Slides not approaching the asymptote of the rarefaction curve were then re-sampled, and analysed iteratively for up to a total of seven transects *APPENDIX 2*.

@@

To estimate the duration of dates in which plant species were flowering weibull estimates of several phenological parameters all spatially modelled taxa were developed (@belitz2020accuracy, @pearse2017statistical). Only BIEN records which occurred in the Omernik Level 4 Ecoregions within 15km of the study area (n = 5 Level 4 Ecoregions, or conditionally 6 if enough records not be found in the nearest 5), and which were from herbarium records were included. To remove temporally irrelevant herbarium records, i.e. material collected during times which flowering is impossible at the study area due to snow cover, we used the SnowUS dataset (@iler2021conceptual, @tran2019cloud) from 2000-2017 was analyzed for the first three days of contiguous snow absence, and the first three days of contiguous snow cover in Fall. Herbarium records after the 3rd quantile for melt, and the 1st quantile for snow cover of these metrics were removed. Species with > 10 records had their weibull distributions generated for the date when 10% of individuals had begun flowering, when 50% were flowering, and when 90% of individuals had flowered.

@@

Across the entire spatial domain of modelling all ensembled models (n = 968) had an accuracy of 0.84 (95% CI 0.8356 - 0.8443), kappa 0.68, p-value < 0.001, sensitivity = 0.80, specificity = 0.87, AUC = 0.92.

@@

**ACKNOWLEDGMENTS:** Nyree Zerega for assistance obtaining herbaria loans and accessioning our collections at CHIC. Pat Herendeen for assistance with virtually all aspects of preparing pollen vouchers and the identification process. Ian Breckheimer for sharing the SDM predictor variables. Hilary Noble, Zoe Diaz-Martinez, Angela McDonnell, & Elena Loke for assistance with genomic library preparation. We thank the curators at the following herbaria for supplying tissue: Ben Legler at Stillinger (ID), Charles (Rick) Williams at Ray J. Davis (IDS), (B)Ernie Nelson at Rocky Mountain (RM). We thank the original collectors of these specimens: D. Knoke, L. Brummer, J. Boyd, C. Davidson, I. Gilman, M. Kirkpatrick, S. McCauley, J. Smith, K. Taylor, & C. Williams. David Giblin for sharing relevant sections of an advanced draft of FNA V. 15. The Bureau of Land Management is thanked as many plant specimens were collected by R.C.B as a partner or contractor to the agency, Sarah Burnett and Lauren Price are thanked for sharing AIM data. Sanda and NEB are gratefully acknowledged for technical support and generously sharing samples.

@@

*Generalised Linear Models (GLM)*

*Generalised Additive Models (GAM)*

Ensemble learning utilizes many sets of trees, each composed of many decisions, to create a single model. Each independent variable ( - or *feature*, may) become a node on the tree, a location on the tree where a binary decision will move towards a predicted outcome. Each of the decision tree models which ensemble learning utilizes is a weak models, each of which may suffer due to high variance or bias, but which produce better outcomes than would be expected via random chance. When ensembled these models generate a strong model, a model which should have more appropriately balanced variance and bias and predicts outcomes which are more strongly correlated with the expected values than the individual weak models.

*Random Forest (RF)* the training data are continually bootstrap re-sampled, in combination with random subsets of features, to create nodes which attempt to optimally predict a known outcome. A large number of trees are then aggregated, via the most common predictions, to generate a final classification prediction tree. Each individual prediction tree is generated independently of the others.

*Boosted Regression Tree (BRT)* An initial tree is grown, and all other trees are derived sequentially from it, as each new tree is grown the errors in responses from the last tree are weighed more heavily so that the model focuses on selecting dependent variables which refine predictions. All response data and predictor variables are kept available to all trees.
